\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2020-12-09 paper.tex
lights v5

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cM}{\mathcal M}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\cQ}{\mathcal Q}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

@@

\paragraph*{Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.\]
Let us introduce the following useful function
\begin{equation}
  \label{eq:g-tilde-1-2}
  \left\{
    \begin{array}{ll}
        \tilde g_{s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}\\
        \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l)
    \end{array}
  \right.
\end{equation}
and present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$) with every increase of $N$ (see Algorithm~\ref{alg:QNMCEM}).

Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align}
  \label{eq:R-def}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\ \nonumber
  & - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k^{(w+1)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\ \nonumber
  & + (y_i \odot \Phi_i^{(w+1)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\ \nonumber
  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w+1)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
\end{align}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{t_j, \gamma_k^{(w+1)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}. Since $R^{(w)}_{n,k}$ depends on parameters estimates at iteration $w+1$ that have not been computed yet ($\gamma_k^{(w+1)}$, $\lambda_0^{(w+1)}$ and $\phi^{(w+1)}$), we denote $R^{(w^-)}_{n,k}$ the function obtained from~\eqref{eq:R-def} by replacing $(w+1)$ with $(w)$, that is using the last iteration estimates.
Hence, we update $\beta_k^{(w)}$ by solving the following problem
\begin{equation}
  \label{eq:sub-problem-beta_k}
  \begin{split}
    \text{minimize}& \quad \quad R_{n, k}^{(w^-)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 + \tilde \eta \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) \big) \\
    \text{subject to}& \quad \quad \beta_{k,j}^+ \geq 0 \text{ and } \beta_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, q.
  \end{split} 
\end{equation}
The value of the gradient for the Gaussian case is here given by
\begin{align*}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w^-)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \\
  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \\
  &+ U_{il}^\top \phi_l^{-2}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{align*}
for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.

@@

Concerning the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{t_i, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w+1)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ therefore requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{p+L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
We then rewrite problem~\eqref{eq:minimization-beta_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-gamma_k}
  \begin{split}
    \text{minimize}& \quad \quad Q_{n, k}^{(w)}(\gamma_k^+ - \gamma_k^-) + \zeta_{2,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\gamma_k^l}^+ - {\gamma_k^l}^-}_2 + \tilde \eta \sum_{j=1}^{p+L\mathcal{A}} (\gamma_{k,j}^+ + \gamma_{k,j}^-) \big) \\
    \text{subject to}& \quad \quad \gamma_{k,j}^+ \geq 0 \text{ and } \gamma_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p+L\mathcal{A}.
  \end{split} 
\end{equation}
One can also compute the gradient since one has
\begin{equation*}
  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,0}} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i
\end{equation*}
and
\begin{align*}
  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\
   &- \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

@@

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-beta_k} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k} \\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

\subsection{Practical details}
\label{sec:practical details}

Let us give some details about the choices made for using the lights model in practice.

@@

When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:Metrics}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness of fit nor generalisation power in a prognostic prediction perspective~\citep{hickey2016joint}. However, practitioners will naturally require predictive prognostic tools to evaluate model fits and compare models.

The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our QNMCEM algorithm introduced in Section~\ref{sec:QNMCEM}), posterior risk classification can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
We propose to use this probabilities as discriminative marker rule for risk prediction and use it in the cross-validation procedure for selecting the best regularization hyper-parameters. Indeed, it makes sense from a practical point of view, since one wants to obtain prognostic predictions as well as interpretation of the high-dimensional longitudinal features.

\paragraph{The case $K=2$.}

The hyper-parameters $(\zeta_1, \zeta_2, \zeta_3)^\top \in \R_+^3$ are then tuned during a 10-fold randomized search cross-validation procedure~\citep{bergstra2012random} using marker $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}) and the C-index score defined in Section~\ref{sec:Metrics}.
Random search are indeed more efficient in terms of computing times than classical grid search for hyper-parameter optimization and finds better models by effectively searching a larger configuration space, which is appropriate to our high-dimensional problem.
% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2, \zeta_3) en 3D

@@

In this section, we first introduce the C-index metric used to assess performances and briefly introduce the models we consider for performance comparisons. Then, we provide details regarding the simulation study and data generation. The chosen metrics for evaluating performances are then presented, followed by the results.

>>>>

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cM}{\mathcal M}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\cQ}{\mathcal Q}
\newcommand{\cR}{\mathcal R}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

@@

\paragraph*{Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.\]
Let us introduce the following useful functions
\begin{equation}
  \label{eq:g-tilde-1-2}
  \left\{
    \begin{array}{ll}
        \tilde g_{s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}, \\
        \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l).
    \end{array}
  \right. 
\end{equation}
We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:QNMCEM}) with every increase of $N$.

Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
  & + (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\
  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
Hence, we update $\beta_k^{(w)}$ by solving the following problem
\begin{equation}
  \label{eq:sub-problem-beta_k}
  \begin{split}
    \text{minimize}& \quad \quad R_{n, k}^{(w)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 + \tilde \eta \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) \big) \\
    \text{subject to}& \quad \quad \beta_{k,j}^+ \geq 0 \text{ and } \beta_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, q.
  \end{split} 
\end{equation}
The value of the gradient for the Gaussian case is here given by
\begin{align}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \\ \nonumber
  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \\ \nonumber
  &+ U_{il}^\top {\phi_l^{(w)}}^{-2}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{align}
for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.

Concerning the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{t_i, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}. The update for $\gamma_k^{(w)}$ therefore requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{p+L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
We then rewrite problem~\eqref{eq:minimization-beta_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-gamma_k}
  \begin{split}
    \text{minimize}& \quad \quad Q_{n, k}^{(w)}(\gamma_k^+ - \gamma_k^-) + \zeta_{2,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\gamma_k^l}^+ - {\gamma_k^l}^-}_2 + \tilde \eta \sum_{j=1}^{p+L\mathcal{A}} (\gamma_{k,j}^+ + \gamma_{k,j}^-) \big) \\
    \text{subject to}& \quad \quad \gamma_{k,j}^+ \geq 0 \text{ and } \gamma_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p+L\mathcal{A}.
  \end{split} 
\end{equation}
One can also compute the gradient since one has
\begin{equation}
  \label{eq:grad-Q-1}
  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,0}} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i
\end{equation}
and
\begin{align}
  \label{eq:grad-Q-2}
  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

@@

@@

\begin{algorithm}[]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-beta_k} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k} \\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

\subsection{The case $K=2$}
\label{sec:K=2}

@@

The hyper-parameters $(\zeta_1, \zeta_2, \zeta_3)^\top \in \R_+^3$ are then tuned during a 10-fold randomized search cross-validation procedure~\citep{bergstra2012random} using predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}.
Random search is indeed more efficient in terms of computing times than classical grid search for hyper-parameters optimization and finds better models by effectively searching in a larger configuration space, which is appropriate to our high-dimensional problem. Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:numerical-details}, while the search intervals for $\zeta_2$ and $\zeta_3$ are determined empirically.
% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2, \zeta_3) en 3D

@@

In this section, we first present the proposed evaluation strategy to assess real-time prognostic prediction performances. We then introduce the appropriate metric used to this end: the C-index ; and finally, the considered models for performance comparisons.

\subsection{Evaluation strategy: the real-time prediction paradigm}
\label{sec:evaluation strategy}

When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:Metrics}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness-of-fit nor generalization power in a prognostic prediction perspective~\citep{hickey2016joint}. However, with the prospect of using prognostic prediction in real-time on a daily basis, practitioners will naturally require predictive prognostic tools to evaluate models fit and compare models in terms of discriminative power. So we choose to put ourselves in this paradigm.

\paragraph{Predictive marker.}
